---
title: "LASSO Regression Report - Data Science Capstone Project"
author: "Nikhitha Amireddy"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

In the complex domain of data science and statistical analysis, the journey to decipher vast datasets has led to the adoption of innovative methodologies, among which LASSO regression shines as a key analytical tool. Esteemed for its ability to streamline extensive data into digestible and actionable insights, LASSO regression serves as a cornerstone for researchers tackling the dynamic interplay of variables across socioeconomic landscapes, health outcomes, and labor market behaviors. This study, grounded in the seminal contributions of Efron et al. (2004) on Least Angle Regression and complemented by the practical insights from Sohaee and Bohluli (2024a; 2024b), aims to leverage the strengths of LASSO regression. We explore its efficacy in discerning the nuanced factors that influence categorical outcomes, from the socioeconomic drivers behind fatal traffic accidents to the determinants of breast cancer screening adherence. Our approach transcends the traditional use of LASSO regression as a mere computational mechanism, employing it instead as a profound lens to understand the intricate web of relations that mold our societal fabric.
As the landscape of data-driven decision-making evolves, the urgency to extract meaningful insights from the plethora of available data intensifies. Inspired by the pioneering work of Robert Tibshirani (1996), our analysis delves deep into the relationships among socioeconomic, educational, and demographic variables, utilizing a rich dataset that spans wage information, educational backgrounds, and demographic specifics. Through diligent application of LASSO regression, we seek to uncover the subtle yet significant connections and predictors affecting wage earnings and educational attainment. This exploration not only demonstrates LASSO regression's adeptness at managing high-dimensional data but also highlights its integral role in illuminating the complex dynamics at the intersection of education, socioeconomic status, and labor market performance.

## Literature Review

The advent of LASSO regression by Robert Tibshirani (1996) heralded a paradigm shift in statistical modeling, introducing a robust mechanism capable of surmounting the enduring challenges of feature selection and model overfitting. This methodology, with its dual functionality for variable selection and regularization, revolutionized the interpretability and precision of predictive models, particularly for high-dimensional datasets. It has since become a beacon for researchers navigating the intricate landscapes of data.
The versatility of LASSO regression has been demonstrated across a multitude of domains, illustrating its capacity to disentangle complex data relationships. The work of Sohaee and Bohluli (2024a; 2024b) in public health and safety research exemplifies LASSO's potential, uncovering critical socioeconomic and demographic factors impacting public health and safety outcomes. Their findings underscore the method's ability to identify pivotal influences, providing essential insights for policy development and strategic interventions.
Bridging the gap between theoretical foundations and practical applications, LASSO regression has undergone continuous innovation and adaptation. The foundational theories proposed by Efron et al. (2004) on Least Angle Regression have enriched our understanding of LASSO's applicability to high-dimensional data analysis, facilitating its integration into real-world research. This progression underscores the significance of LASSO regression in addressing contemporary societal challenges.
Yet, the journey of LASSO regression is marked by ongoing exploration and refinement. Challenges such as tuning parameter selection, interpretation amidst correlated predictors, and methodological assumptions call for further research. Future directions suggest the development of more adaptive algorithms, integration with emerging machine learning techniques, and exploration of LASSO's applicability in new research areas, promising to enhance LASSO's utility and effectiveness in statistical analysis further.


## Methods

The common non-parametric regression model is
$Y_i = m(X_i) + \varepsilon_i$, where $Y_i$ can be defined as the sum of
the regression function value $m(x)$ for $X_i$. Here $m(x)$ is unknown
and $\varepsilon_i$ some errors. With the help of this definition, we
can create the estimation for local averaging i.e. $m(x)$ can be
estimated with the product of $Y_i$ average and $X_i$ is near to $x$. In
other words, this means that we are discovering the line through the
data points with the help of surrounding data points. The estimation
formula is printed below [@R-base]:

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$ $W_n(x)$ is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if $X_i$ is far from $x$.

Another equation:

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data and Visualization

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Statistical Modeling

```{r}

```

### Conclusion

## References

[1] Efron, B., Hastie, T., Johnstone, I., & Tibshirani, R. (2004). "Least angle regression" (Annals of Statistics, 32(2), 407-499).

[2] Sohaee, R., & Bohluli, M. (2024a). "Effect of Neighborhood and Individual-Level Socioeconomic Factors on Breast Cancer Screening Adherence" (Journal of Public Health).

[3] Sohaee, R., & Bohluli, M. (2024b). "Nonlinear Analysis of the Effects of Socioeconomic, Demographic, and Technological Factors on the Number of Fatal Traffic Accidents" (Safety, 10(1)).

[4] Tibshirani, R. (1996). "Regression shrinkage and selection via the lasso" (Journal of the Royal Statistical Society: Series B (Methodological), 58(1), 267-288).

