---
title: "LASSO Regression Report - Data Science Capstone Project"
author: "Nikhitha Amireddy, Ishrath Jahan, Sai Kumar Miryala, Muhammad Usman Aslam"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

The advent of the Least Absolute Shrinkage and Selection Operator (LASSO) regression marks a significant milestone in the field of statistical analysis and predictive modeling. Introduced by Tibshirani (1996) [@RegressionShrinkage], this methodology has revolutionized the way researchers approach large, complex datasets, offering a sophisticated tool for variable selection and regularization. LASSO regression's ability to penalize the absolute size of regression coefficients effectively reduces model complexity, enhancing interpretability without sacrificing predictive power. This dual capacity to streamline data analysis and mitigate overfitting challenges has rendered LASSO an invaluable asset across various scientific domains, from public health to economic forecasting and environmental studies.

The proliferation of data in recent years has underscored the critical need for methodologies that can not only manage but also extract meaningful insights from high-dimensional datasets. LASSO regression, with its foundation in the work of Efron et al. (2004) [@annalsofstatistics] on Least Angle Regression, has emerged as a key analytical tool in deciphering the intricate interplay of variables that characterize complex socioeconomic landscapes, health outcomes, and labor market behaviors. Through the lens of LASSO regression, researchers have been able to uncover nuanced factors influencing categorical outcomes, such as the socioeconomic drivers behind fatal traffic accidents and the determinants of breast cancer screening adherence (Sohaee & Bohluli, 2024a; 2024b)[@safety10010011], [@38166942]. This exploration transcends traditional computational applications, employing LASSO regression as a profound instrument for understanding the multifaceted relationships that shape societal dynamics.

Inspired by the pioneering contributions of Tibshirani (1996)[[@RegressionShrinkage], this collective body of work endeavors to harness the strengths of LASSO regression in revealing the relationships among socioeconomic, educational, and demographic variables. Utilizing a rich dataset encompassing wage information, educational backgrounds, and demographic specifics, our analysis aims to illuminate the complex dynamics at play within the realms of education, socioeconomic status, and labor market performance. Such an endeavor not only showcases LASSO regression's adeptness at handling high-dimensional data but also emphasizes its critical role in demystifying the intricate connections and predictors affecting wage earnings and educational attainment.

## Literature Review

The introduction of LASSO regression by Tibshirani (1996) [@RegressionShrinkage] represented a paradigm shift in statistical modeling, addressing long-standing challenges associated with feature selection and model overfitting. This method's elegant simplicity, offering a robust mechanism for both variable selection and regularization, has significantly improved the interpretability and precision of predictive models, particularly when dealing with complex, high-dimensional datasets. The foundational work of Efron et al. (2004) further elucidated the methodological underpinnings of LASSO, demonstrating its versatility and efficiency through the concept of Least Angle Regression.

The applicability of LASSO regression spans a diverse array of fields, evidencing its capacity to untangle the web of complex data relationships. In the domain of public health, the studies by Sohaee and Bohluli (2024a; 2024b) [@safety10010011], [@38166942] serve as exemplars of LASSO's utility in identifying critical socioeconomic and demographic factors impacting health outcomes and safety. These investigations underscore the method's adeptness in pinpointing influential variables, thereby facilitating the formulation of targeted policies and interventions.

The economic sector has also benefitted from the application of LASSO regression, where it has been employed to dissect the determinants of economic development and labor market fluctuations. Research into the socioeconomic factors influencing high-quality economic growth in China, for instance, leverages LASSO to parse through vast datasets, revealing insights into the drivers of economic resilience and advancement (Chunli Zhoua & X. L., 2022)[@ZHOU20221055]. Such analyses not only affirm LASSO's capability in managing multidimensional data but also highlight its significance in informing economic policy and strategy.

Environmental studies, particularly those concerning operational efficiency and sustainability, have similarly utilized LASSO regression to forecast ship fuel consumption and optimize maritime logistics. The work of Wang et al. (2017)[@WANG2018817] exemplifies the algorithm's strength in addressing challenges related to variable correlation and model interpretability, offering a path toward more sustainable and efficient operational practices.

Despite its widespread application, the journey of LASSO regression is characterized by continuous methodological innovation and adaptation. The evolution from its initial proposition to its current state reflects a dynamic interplay between theoretical advancements and practical implementations. Challenges such as the selection of tuning parameters, interpretation in the presence of correlated predictors, and the overarching assumptions underlying the LASSO model invite ongoing scholarly inquiry.

Future research directions point toward the development of more adaptive LASSO algorithms, the integration of LASSO with cutting-edge machine learning techniques, and the exploration of its utility in novel research areas. Such endeavors promise not only to extend the boundaries of LASSO's applicability but also to enhance its efficacy in statistical analysis and predictive modeling.



## Methods

The common non-parametric regression model is
$Y_i = m(X_i) + \varepsilon_i$, where $Y_i$ can be defined as the sum of
the regression function value $m(x)$ for $X_i$. Here $m(x)$ is unknown
and $\varepsilon_i$ some errors. With the help of this definition, we
can create the estimation for local averaging i.e. $m(x)$ can be
estimated with the product of $Y_i$ average and $X_i$ is near to $x$. In
other words, this means that we are discovering the line through the
data points with the help of surrounding data points. The estimation
formula is printed below [@R-base]:

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$ $W_n(x)$ is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if $X_i$ is far from $x$.

Another equation:

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data and Visualization

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Statistical Modeling

```{r}

```

### Conclusion

## References


[1]	Chunli Zhoua, X. L. (2022). "Influencing factors of the high-quality economic development in China based on LASSO model." Energy Reports, 8, 1055–1065.
[2]	Efron, B., Hastie, T., Johnstone, I., & Tibshirani, R. (2004). "Least angle regression." Annals of Statistics, 32(2), 407-499.
[3]	Sohaee, R., & Bohluli, M. (2024a). "Effect of Neighborhood and Individual-Level Socioeconomic Factors on Breast Cancer Screening Adherence." Journal of Public Health.
[4]	Sohaee, R., & Bohluli, M. (2024b). "Nonlinear Analysis of the Effects of Socioeconomic, Demographic, and Technological Factors on the Number of Fatal Traffic Accidents." Safety, 10(1).
[5]	Tibshirani, R. (1996). "Regression shrinkage and selection via the lasso." Journal of the Royal Statistical Society: Series B (Methodological), 58(1), 267-288.
[6]	Wang, S. J., et al. (2017). "Predicting ship fuel consumption based on LASSO regression." Elsevier, 1361-9209.


